---
title: "SLiM_PFE"
author: "Sam Bogan"
date: "2025-01-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggpubr)
library(gridExtra)
library(patchwork)
library(png)
library(grid)

```

Figure 1 diagram

```{r}

# Generate conceptual data frame
reaction_norm_data <- data.frame(Scenario = c("Genotype", "Genotype", 
                                              "Genotype", "Genotype", 
                                              "Genotype", "Genotype",
                                              "Genotype + environment", "Genotype + environment", 
                                              "Genotype + environment", "Genotype + environment", 
                                              "Genotype + environment", "Genotype + environment",
                                              "Genotype x environment", "Genotype x environment", 
                                              "Genotype x environment", "Genotype x environment", 
                                              "Genotype x environment", "Genotype x environment"),
                                 Environment = c(1,2,1,2,1,2,
                                                 1,2,1,2,1,2,
                                                 1,2,1,2,1,2),
                                 Genotype = c(1,1,2,2,3,3,
                                              1,1,2,2,3,3,
                                              1,1,2,2,3,3),
                                 Phenotype = c(1,1,2,2,3,3,
                                               1,2,2,3,3,4,
                                               3,3,2,3,1,3),
                                 RN = c(0,0,0,0,0,0,
                                        1,1,1,1,1,1,
                                        0,0,1,1,2,2),
                                 Int = c(1,1,2,2,3,3,
                                         1,1,2,2,3,3,
                                         3,3,2,2,1,1))

# Create panel A
reaction_norm_plot <- ggplot(reaction_norm_data, aes(x = Environment, y = Phenotype, group = Genotype)) +
  geom_line(aes(color = as.factor(RN)), size = 2) +
  geom_point(data = filter(reaction_norm_data, Environment < 2),
             aes(shape = as.character(Phenotype)), size = 6, color = "black", show.legend = TRUE) +
  geom_point(data = filter(reaction_norm_data, Environment < 2),
             aes(shape = as.character(Phenotype)), color = "white", size = 4, show.legend = TRUE) +
  facet_wrap(~Scenario, nrow = 1, ncol = 3) +
  labs(x = "Environment", y = "Phenotype", color = "Reaction \nnorm", fill = "", shape = "Phenotypic \nintercept") +
  theme_classic(base_size = 20) +
  theme(strip.background = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "right",
        legend.direction = "vertical",
        legend.box.spacing = unit(2.2, "cm")) +
  scale_y_continuous(expand = expansion(mult = 0.2), limits = c(1,4)) +
  scale_color_manual(values = c("black", "gray40", "gray80")) +
  guides(color = guide_legend(override.aes = list(
    shape = NA,      # no point shape
    linetype = 1,    # solid line
    size = 2         # match line size
)))

reaction_norm_plot

```

```{r}

library(tibble)
library(ggforce)


# Create base structure with adjusted coordinates
pb_df <- tibble::tribble(
  ~Var, ~Arch, ~Group, ~Shape, ~Fill, ~Size, ~x, ~y, ~alpha,
  
  # Monogenic (centered and evenly spaced)
  "Genotype", "Monogenic \n(1 gene)", NA, "3", "0", 3, 0, 3, 1,
  "Genotype", "Monogenic \n(1 gene)", NA, "1", "0", 3, 0, 1, 1,
  "Genotype", "Monogenic \n(1 gene)", NA, "2", "0", 3, 0, 2, 1,

  "Genotype + environment", "Monogenic \n(1 gene)", NA, "3", "1", 3, 1, 3, 1,
  "Genotype + environment", "Monogenic \n(1 gene)", NA, "1", "1", 3, 1, 1, 1,
  "Genotype + environment", "Monogenic \n(1 gene)", NA, "2", "1", 3, 1, 2, 1,

  "Genotype x environment", "Monogenic \n(1 gene)", NA, "3", "0", 3, 0, 3, 1,
  "Genotype x environment", "Monogenic \n(1 gene)", NA, "1", "2", 3, 2, 1, 1,
  "Genotype x environment", "Monogenic \n(1 gene)", NA, "2", "1", 3, 1, 2, 1,
  
  # Epistatic (centered and evenly spaced)
  "Genotype", "Epistatic \n(1 gene)", NA, "3", "0", 3, 0, 3, 1,
  "Genotype", "Epistatic \n(1 gene)", NA, "1", "0", 3, 0, 1, 1,
  "Genotype", "Epistatic \n(1 gene)", NA, "2", "0", 3, 0, 2, 1,

  "Genotype + environment", "Epistatic \n(1 gene)", NA, "3", "1", 3, 1, 3, 1,
  "Genotype + environment", "Epistatic \n(1 gene)", NA, "1", "3", 3, 1, 1, 1,
  "Genotype + environment", "Epistatic \n(1 gene)", NA, "2", "2", 3, 1, 2, 1,

  "Genotype x environment", "Epistatic \n(1 gene)", NA, "3", "0", 3, 0, 3, 1,
  "Genotype x environment", "Epistatic \n(1 gene)", NA, "1", "4", 3, 2, 1, 1,
  "Genotype x environment", "Epistatic \n(1 gene)", NA, "2", "2", 3, 1, 2, 1,
  
  # Polygenic - Arranged into equilateral triangle layout
  # Group centers at (1,2), (2,3.73), (3,2) for triangle vertices
  # Internal points grouped tightly around each center

  # G1
  "Genotype", "Polygenic \n(3 genes)", "G1", "3", "0", 1, 0, 3.15, 1,
  "Genotype", "Polygenic \n(3 genes)", "G1", "3", "0", 1, -.15, 2.85, 1,
  "Genotype", "Polygenic \n(3 genes)", "G1", "3", "0", 1, 0.15, 2.85, 1,

  # G2
  "Genotype", "Polygenic \n(3 genes)", "G2", "3", "0", 1, 0, 2.15, 1,
  "Genotype", "Polygenic \n(3 genes)", "G2", "1", "0", 1, -.15, 1.85, 1,
  "Genotype", "Polygenic \n(3 genes)", "G2", "2", "0", 1, 0.15, 1.85, 1,

  # G3
  "Genotype", "Polygenic \n(3 genes)", "G3", "1", "0", 1, 0, 1.15, 1,
  "Genotype", "Polygenic \n(3 genes)", "G3", "1", "0", 1, -.15, 0.85, 1,
  "Genotype", "Polygenic \n(3 genes)", "G3", "1", "0", 1, 0.15, 0.85, 1,

  # G+E
  "Genotype + environment", "Polygenic \n(3 genes)", "G1", "3", "1", 1, 1, 3.15, 1,
  "Genotype + environment", "Polygenic \n(3 genes)", "G1", "3", "1", 1, 1.15, 2.85, 1,
  "Genotype + environment", "Polygenic \n(3 genes)", "G1", "3", "1", 1, 0.85, 2.85, 1,

  "Genotype + environment", "Polygenic \n(3 genes)", "G2", "3", "1", 1, 1, 2.15, 1,
  "Genotype + environment", "Polygenic \n(3 genes)", "G2", "1", "1", 1, 0.85, 1.85, 1,
  "Genotype + environment", "Polygenic \n(3 genes)", "G2", "2", "1", 1, 1.15, 1.85, 1,

  "Genotype + environment", "Polygenic \n(3 genes)", "G3", "1", "1", 1, 1, 1.15, 1,
  "Genotype + environment", "Polygenic \n(3 genes)", "G3", "1", "1", 1, 1.15, 0.85, 1,
  "Genotype + environment", "Polygenic \n(3 genes)", "G3", "1", "1", 1, 0.85, 0.85, 1,

  # GxE
  "Genotype x environment", "Polygenic \n(3 genes)", "G1", "3", "0", 1, 0, 3.15, 1,
  "Genotype x environment", "Polygenic \n(3 genes)", "G1", "3", "0", 1, -.15, 2.85, 1,
  "Genotype x environment", "Polygenic \n(3 genes)", "G1", "3", "0", 1, 0.15, 2.85, 1,

  "Genotype x environment", "Polygenic \n(3 genes)", "G2", "1", "2", 1, 2.0, 1.15, 1,
  "Genotype x environment", "Polygenic \n(3 genes)", "G2", "1", "2", 1, 1.85, 0.85, 1,
  "Genotype x environment", "Polygenic \n(3 genes)", "G2", "1", "2", 1, 2.15, 0.85, 1,

  "Genotype x environment", "Polygenic \n(3 genes)", "G3", "3", "0", 1, 1.0, 2.15, 1,
  "Genotype x environment", "Polygenic \n(3 genes)", "G3", "1", "2", 1, 0.85, 1.85, 1,
  "Genotype x environment", "Polygenic \n(3 genes)", "G3", "2", "1", 1, 1.15, 1.85, 1
)

# Circle outlines for Polygenic groups
poly_circles <- pb_df %>%
  filter(!is.na(Group)) %>%
  group_by(Var, Arch, Group) %>%
  summarise(x0 = mean(x), y0 = mean(y), r = .4, .groups = "drop")

# Shape map
shape_map <- c("3" = 22, "1" = 21, "2" = 24)

# Set factor order for facets
pb_df$Arch <- factor(pb_df$Arch, levels = c("Monogenic \n(1 gene)", "Polygenic \n(3 genes)", "Epistatic \n(1 gene)"))
poly_circles$Arch <- factor(poly_circles$Arch, levels = c("Monogenic \n(1 gene)", "Polygenic \n(3 genes)", "Epistatic \n(1 gene)"))

# Plot
pb_df$Fill <- factor(pb_df$Fill, levels = c("0", "1", "2", "3", "4"))
pb_df$Shape <- factor(pb_df$Shape, levels = c("1", "2", "3"))

Fig1B_rev <- ggplot(pb_df, aes(x = x, y = y, fill = Fill)) +
  geom_circle(data = poly_circles, aes(x0 = x0, y0 = y0, r = r),
              inherit.aes = FALSE, color = "black", linetype = "dotted", size = .75) +
  geom_point(aes(shape = Shape), size = 6, color = "black", stroke = 1.2) +
  scale_shape_manual(values = shape_map) +
  scale_fill_manual(
    values = c("0" = "black", "1" = "gray40", "2" = "gray80", "3" = "gray95", "4" = "white"),
    drop = FALSE
  ) +
  facet_grid(Arch ~ Var) +
  labs(shape = "Phenotypic \nintercept \neffect", fill = "Reaction \nnorm \neffect", size = "Allele \neffect \nstrength",
       x = "Reaction norm", y = "Phenotypic intercept") +
  theme_classic(base_size = 20) +
  scale_size_manual(values = c(3,6)) +
  theme(
    strip.text.y = element_text(angle = 270),
    legend.position = "right",
    legend.direction = "vertical",
    strip.background = element_blank(),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.5)) +
  guides(fill = guide_legend(override.aes = list(
    shape = 21,
    size = 4.5,
    linetype = 1)),
    shape = guide_legend(override.aes = list(
    size = 4.5))) +
  scale_x_continuous(breaks = c(0, 1, 2), limits = c(-.8, 2.7)) +
  scale_y_continuous(breaks = c(1, 2, 3), limits = c(0.3, 3.7))

Fig1B_rev

```

```{r}

# Arrange revised Figure 1A and 1B panels
Fig1_rev <- ggarrange(reaction_norm_plot, Fig1B_rev,
                      labels = c("A", "B"),
                      heights = c(.32,.68),
                      ncol = 1, nrow = 2,
                      align = "hv")

Fig1_rev

# Save figure
png("~/Documents/GitHub/PFE_perspective/Fig1_rev.png",
    units = "in", width = 12, height = 14, res = 600)
Fig1_rev
dev.off()

```

```{r}
# Save figure
png("~/Documents/GitHub/PFE_perspective/Fig1A_rev.png", 
    units = "in", width = 6, height = 12, res = 600)
reaction_norm_plot
dev.off()

```

```{r}
# Simulate data for evolutionary change
n <- 1000

# Plasticity-first evolution: High rate of phenotypic change
plasticity_first <- data.frame(
  Plasticity = rnorm(n, mean = 2, sd = 0.5),
  Phenotype = pmax(rnorm(n, mean = 4, sd = 0.5), 0),
  Selection = rnorm(n, mean = 0.2, sd = 0.05),
  Scenario = "Plasticity-first"
)

# Canalization: High rate of phenotypic change
canalization <- data.frame(
  Plasticity = rnorm(n, mean = -2, sd = 0.5),
  Phenotype = pmax(rnorm(n, mean = 4, sd = 0.5), 0),
  Selection = rnorm(n, mean = -0.2, sd = 0.05),
  Scenario = "Canalization"
)

# Plasticity-buffered evolution: Low rate of phenotypic change near 0
plasticity_buffered <- data.frame(
  Plasticity = rnorm(n, mean = 1, sd = 0.5),
  Phenotype = pmax(rnorm(n, mean = 0.5, sd = 0.1), 0),
  Selection = rnorm(n, mean = 0.2, sd = 0.05),
  Scenario = "Plasticity-buffered"
)

# Combine data
evolution_data <- rbind(plasticity_first, canalization, plasticity_buffered)

# Plot histograms for plasticity and selection coefficients
plasticity_histogram <- ggplot(evolution_data, aes(x = Plasticity)) +
  geom_density(alpha = 1, fill = "gray", position = "identity") +
  facet_wrap(~ Scenario, scales = "free_y") +
  labs(title = "Rates of Evolutionary Change: Plasticity", x = "Plasticity", y = "Frequency") +
  theme_classic(base_size = 20) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "none")

phenotype_histogram <- ggplot(evolution_data, aes(x = Phenotype)) +
  geom_density(alpha = 1, fill = "gray", position = "identity") +
  facet_wrap(~ Scenario, scales = "free_y") +
  labs(title = "Rate of phenotypic evolution", x = "Evolutionary rate", y = "Frequency") +
  theme_classic(base_size = 20) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "none")

selection_histogram <- ggplot(evolution_data, aes(x = Selection)) +
  geom_density(alpha = 1, fill = "gray", position = "identity") +
  facet_wrap(~ Scenario, scales = "free_y") +
  labs(title = "Selection on plasticity", x = "Selection coefficient", y = "Frequency") +
  theme_classic(base_size = 20) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "none")

# Combine evolutionary plots
evolution_plots <- ggarrange(phenotype_histogram, selection_histogram, 
                             labels = c("", ""), ncol = 1, nrow = 2)

# --- Combine All Plots ---
Fig1_rev <- ggarrange(reaction_norm_plot, evolution_plots, 
                      labels = c("A", "B"),
                      heights = c(.5, 1),
                      ncol = 1, nrow = 2)

# Display final plot
print(Fig1_rev)

# Export the comparison plot as PNG
png("~/Documents/GitHub/PFE_perspective/Fig1_rev.png", 
    units = "in", width = 12, height = 12, res = 600)
Fig1_rev
dev.off()

```

Figure 2 simulation of epistatic PFE

```{r}

# Parameters
set.seed(42)
generations <- 100
population_size <- 1000
recomb_rate <- 0.02
drift_sd <- 0.005
temperature <- seq(20, 35, length.out = generations) + rnorm(generations, sd = 0.5)

# Initial haplotype frequencies
freqs <- data.frame(
  generation = 1:generations,
  axbx = 0.0,   # Completely purged initially
  axby = 0.10,  # Initial frequency
  other = 0.90  # Other haplotypes
)

# Selection coefficients for scenarios
selection_scenario1 <- function(temp) {
  list(
    axbx = ifelse(temp > 30, 0.15, ifelse(temp > 25, -0.05, -0.15)),
    axby = ifelse(temp > 18 & temp < 28, 0.05, ifelse(temp > 28, -0.05, 0))
  )
}
selection_scenario2 <- function(temp) {
  list(
    axbx = ifelse(temp > 30, 0.15, ifelse(temp > 25, -0.05, -0.15)),
    axby = 0  # Neutral under all conditions
  )
}

# Simulation function
simulate_haplotypes <- function(selection_function) {
  results <- freqs
  recombination_events <- numeric(generations)
  
  for (t in 2:generations) {
    temp <- temperature[t]
    sel <- selection_function(temp)
    
    # Recombination: Stronger dependence on axby and bx alleles
    recombination_events[t] <- recomb_rate * results$axby[t-1] * results$other[t-1]
    new_axbx <- recombination_events[t]
    
    # Selection and drift
    axbx_fit <- results$axbx[t-1] + sel$axbx * results$axbx[t-1] + new_axbx
    axby_fit <- results$axby[t-1] + sel$axby * results$axby[t-1] - new_axbx
    
    # Drift
    axbx_fit <- axbx_fit + rnorm(1, mean = 0, sd = drift_sd)
    axby_fit <- axby_fit + rnorm(1, mean = 0, sd = drift_sd)
    
    # Normalize frequencies
    other_fit <- max(0, 1 - axbx_fit - axby_fit)
    total <- axbx_fit + axby_fit + other_fit
    
    results$axbx[t] <- max(0, axbx_fit / total)
    results$axby[t] <- max(0, axby_fit / total)
    results$other[t] <- max(0, other_fit / total)
  }
  
  results$recombination <- recombination_events
  results
}

# Run simulations for both scenarios
scenario1 <- simulate_haplotypes(selection_scenario1)
scenario1$scenario <- "PFE"

scenario2 <- simulate_haplotypes(selection_scenario2)
scenario2$scenario <- "Neutral plasticity"

# Update for relabeling and adding selection coefficients over time
scenario1 <- scenario1 %>%
  mutate(sel_axbx = map_dbl(1:generations, ~ selection_scenario1(temperature[.x])$axbx),
         sel_axby = map_dbl(1:generations, ~ selection_scenario1(temperature[.x])$axby))

scenario2 <- scenario2 %>%
  mutate(sel_axbx = map_dbl(1:generations, ~ selection_scenario2(temperature[.x])$axbx),
         sel_axby = map_dbl(1:generations, ~ selection_scenario2(temperature[.x])$axby))

# Combine results
results <- bind_rows(scenario1, scenario2) %>%
  pivot_longer(cols = c(axbx, axby, other), names_to = "haplotype", values_to = "frequency")

# Separate datasets for clarity
selection_results <- results %>%
  filter(haplotype == "axbx") %>%
  select(generation, scenario, sel_axbx, sel_axby) %>%
  distinct()

recombination_results <- results %>%
  filter(haplotype == "axbx") %>%
  select(generation, scenario, recombination) %>%
  distinct()

temperature_results <- data.frame(
  generation = 1:generations,
  temperature = temperature,
  scenario = rep(c("PFE", "Neutral plasticity"), each = generations)
)

```

Simulation with logistic selection

```{r}

# Parameters
set.seed(42)
generations <- 100
population_size <- 1000
recomb_rate <- 0.02
drift_sd <- 0.005
temperature <- seq(20, 35, length.out = generations) + rnorm(generations, sd = 0.5)

# Initial haplotype frequencies
freqs <- data.frame(
  generation = 1:generations,
  axbx = 0.0,
  axby = 0.10,
  other = 0.90
)

# Logistic function
logistic_trend <- function(t, midpoint, steepness, start_val, end_val) {
  start_val + (end_val - start_val) / (1 + exp(-steepness * (t - midpoint)))
}

# Shared logistic curve for axbx in both scenarios, delayed increase
shared_axbx_curve <- function(t) {
  logistic_trend(t, midpoint = 60, steepness = 0.11, start_val = -0.15, end_val = 0.15)
}

# Selection coefficients
selection_scenario1 <- function(t) {
  list(
    axbx = shared_axbx_curve(t),
    axby = logistic_trend(t, midpoint = 60, steepness = 0.44, start_val = 0.045, end_val = -0.045)
  )
}

selection_scenario2 <- function(t) {
  list(
    axbx = shared_axbx_curve(t),
    axby = 0
  )
}

selection_scenario3 <- function(t) {
  list(
    axbx = shared_axbx_curve(t),
    axby = logistic_trend(t, midpoint = 60, steepness = 0.055, start_val = 0.2, end_val = -0.045)
  )
}

# Simulation function
simulate_haplotypes <- function(selection_function) {
  results <- freqs
  recombination_events <- numeric(generations)

  for (t in 2:generations) {
    sel <- selection_function(t)

    recombination_events[t] <- recomb_rate * results$axby[t-1] * results$other[t-1]
    new_axbx <- recombination_events[t]

    axbx_fit <- results$axbx[t-1] + sel$axbx * results$axbx[t-1] + new_axbx
    axby_fit <- results$axby[t-1] + sel$axby * results$axby[t-1] - new_axbx

    axbx_fit <- axbx_fit + rnorm(1, mean = 0, sd = drift_sd)
    axby_fit <- axby_fit + rnorm(1, mean = 0, sd = drift_sd)

    other_fit <- max(0, 1 - axbx_fit - axby_fit)
    total <- axbx_fit + axby_fit + other_fit

    results$axbx[t] <- max(0, axbx_fit / total)
    results$axby[t] <- max(0, axby_fit / total)
    results$other[t] <- max(0, other_fit / total)
  }

  results$recombination <- recombination_events
  results
}

# Run simulations
scenario1 <- simulate_haplotypes(selection_scenario1)
scenario1$scenario <- "Moderately adaptive plasticity"

scenario2 <- simulate_haplotypes(selection_scenario2)
scenario2$scenario <- "Neutral plasticity"

scenario3 <- simulate_haplotypes(selection_scenario3)
scenario3$scenario <- "Strongly adaptive plasticity"

# Annotate selection and recombination
scenario1 <- scenario1 %>%
  mutate(sel_axbx = map_dbl(1:generations, ~ selection_scenario1(.x)$axbx),
         sel_axby = map_dbl(1:generations, ~ selection_scenario1(.x)$axby))

scenario2 <- scenario2 %>%
  mutate(sel_axbx = map_dbl(1:generations, ~ selection_scenario2(.x)$axbx),
         sel_axby = map_dbl(1:generations, ~ selection_scenario2(.x)$axby))

scenario3 <- scenario3 %>%
  mutate(sel_axbx = map_dbl(1:generations, ~ selection_scenario3(.x)$axbx),
         sel_axby = map_dbl(1:generations, ~ selection_scenario3(.x)$axby))

# Combine results
results <- bind_rows(scenario1, scenario2, scenario3) %>%
  pivot_longer(cols = c(axbx, axby, other), names_to = "haplotype", values_to = "frequency")

# Additional outputs
selection_results <- results %>%
  filter(haplotype == "axbx") %>%
  select(generation, scenario, sel_axbx, sel_axby) %>%
  distinct()

recombination_results <- results %>%
  filter(haplotype == "axbx") %>%
  select(generation, scenario, recombination) %>%
  distinct()

temperature_results <- data.frame(
  generation = 1:generations,
  temperature = temperature,
  scenario = rep(c("Strongly adaptive plasticity", "Moderately adaptive plasticity", 
                   "Neutral plasticity"), each = generations)
)

```


```{r}

# Plotting
library(patchwork)

# Temperature over time
plot_temp <- ggplot(temperature_results, aes(x = generation, y = temperature)) +
  geom_line(size = 1.5) +
  facet_wrap(~scenario, nrow = 2) +
  theme_classic(base_size = 20) +
  theme(strip.background = element_blank()) +
  labs(y = "Environment", x = "Generation") +
  theme(legend.position = "none")

# Recombination events leading to axbx
recombination_results$scenario <- factor(recombination_results$scenario, 
                                         levels = c("Neutral plasticity", 
                                                    "Moderately adaptive plasticity",
                                                    "Strongly adaptive plasticity"))

plot_recomb <- ggplot(recombination_results, aes(x = generation, y = recombination)) +
  geom_line(size = 1.5) +
  facet_wrap(~scenario, nrow = 3) +
  theme_classic(base_size = 20) +
  theme(strip.background = element_blank()) +
  labs(y = "Recombination generating axbx", x = "Generation") +
  theme(legend.position = "none")

# Frequencies of haplotypes
results$scenario <- factor(results$scenario,
                           levels = c("Neutral plasticity",
                                      "Moderately adaptive plasticity",
                                      "Strongly adaptive plasticity"))

results$haplotype <- ifelse(results$haplotype == "axby", "axby (plastic)",
                            ifelse(results$haplotype == "axbx", "axbx (fixed)", "Other"))

results$Focal <- ifelse(results$haplotype == "Other","Other", "Focal")

plot_haplotypes <- ggplot(filter(results, haplotype %in% c("axby (plastic)", "axbx (fixed)")), 
                          aes(x = generation, y = frequency, color = haplotype)) +
  geom_line(size = 1.5) +
  facet_wrap(~scenario, nrow = 3) +
  theme_classic(base_size = 20) +
  labs(y = "Frequency", x = "Generation", color = "Haplotype") +
  scale_color_manual(values = c("axbx (fixed)" = "black", "axby (plastic)" = "gray80")) +
  theme(strip.background = element_blank(),
        legend.position = "top")

plot_haplotypes

# Selection coefficients
selection_results$scenario <- factor(selection_results$scenario,
                                     levels = c("Neutral plasticity",
                                                "Moderately adaptive plasticity",
                                                "Strongly adaptive plasticity"))

plot_selection <- ggplot(selection_results, aes(x = generation, color = haplotype)) +
  geom_smooth(aes(y = sel_axbx, color = "axbx (fixed)"), size = 1.5, se = FALSE, span = .1) +
  geom_smooth(aes(y = sel_axby, color = "axby (plastic)"), size = 1.5, se = FALSE, span = .1) +
  facet_wrap(~scenario, nrow = 3) +
  theme_classic(base_size = 20) +
  theme(strip.background = element_blank(),
        legend.position = "top") +
  labs( y = "Selection coefficient", x = "Generation", color = "Haplotype") +
  scale_color_manual(values = c("axbx (fixed)" = "black", "axby (plastic)" = "gray80"))

plot_selection

library(ggpubr)

# Combine plots into a grid with ggarrange
Fig4D <- readPNG("Fig4D_rev_crop.png")
Fig4Dg <- rasterGrob(Fig4D, interpolate = TRUE)

Fig4Dg <- ggplot() +
  annotation_custom(Fig4Dg) +
  ggtitle("Coalescent schematic of \nmoderately adaptive \nplasticity scenario") +
  theme_void(base_size = 20) +
  theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(t = 0, r = 30, b = 0, l = 90))  # leftward nudge

final_plot <- ggarrange(
  plot_selection,
  plot_recomb, 
  plot_haplotypes,
  Fig4Dg,
  ncol = 2, nrow = 2, 
  labels = c("A", "B", "C", "D")
)

final_plot

# Display the combined plot
# Export the comparison plot as PNG
png("~/Documents/GitHub/PFE_perspective/Fig4_rev_log.png", 
    units = "in", width = 12, height = 14, res = 600)
final_plot
dev.off()

```

Simulation without assumptions about mutation rate, recombination rate, or population size

```{r}

setwd("~/Documents/GitHub/PFE_perspective/arg_sim/")

# Load necessary library
library(Biostrings)

# Define parameters
total_length <- 100000  # 100 kb region
num_individuals <- 10   # Population size
num_generations <- 50   # Number of generations
alleles <- c("A", "T", "C", "G")  # Possible nucleotides
gene_length <- total_length / 2   # Length for each haplotype
mutation_rate <- 0.001  # Mutation rate per site

# Function to initialize the population with random haplotypes
initialize_population <- function(n, gene_length) {
  population <- data.frame(
    Hap_A1 = replicate(n, paste(sample(alleles, gene_length, replace = TRUE), collapse = "")),
    Hap_A2 = replicate(n, paste(sample(alleles, gene_length, replace = TRUE), collapse = "")),
    Hap_B1 = replicate(n, paste(sample(alleles, gene_length, replace = TRUE), collapse = "")),
    Hap_B2 = replicate(n, paste(sample(alleles, gene_length, replace = TRUE), collapse = ""))
  )
  return(population)
}

# Function to introduce mutations
mutate_haplotype <- function(haplotype, mutation_rate) {
  seq_split <- unlist(strsplit(haplotype, ""))
  mutate_indices <- which(runif(length(seq_split)) < mutation_rate)
  seq_split[mutate_indices] <- sample(alleles, length(mutate_indices), replace = TRUE)
  return(paste(seq_split, collapse = ""))
}

mutate_population <- function(population, mutation_rate) {
  population[] <- lapply(population, function(hap) sapply(hap, mutate_haplotype, mutation_rate))
  return(population)
}

# Function to perform recombination
recombine <- function(population) {
  n <- nrow(population)
  for (i in seq_len(n)) {
    partner <- sample(seq_len(n), 1)
    if (runif(1) < 0.5) {
      temp <- population$Hap_A2[i]
      population$Hap_A2[i] <- population$Hap_A2[partner]
      population$Hap_A2[partner] <- temp
    }
    if (runif(1) < 0.5) {
      temp <- population$Hap_B2[i]
      population$Hap_B2[i] <- population$Hap_B2[partner]
      population$Hap_B2[partner] <- temp
    }
  }
  return(population)
}

# Function to apply selection while maintaining population size
apply_selection <- function(population, selection_criteria) {
  selected <- switch(
    selection_criteria,
    "plastic" = which(substr(population$Hap_A1, 1, 1) == "A" & substr(population$Hap_B1, 1, 1) == "C"),
    "canalized" = which(substr(population$Hap_A1, 1, 1) == "T" & substr(population$Hap_B1, 1, 1) == "G")
  )
  
  # Ensure selected population is exactly num_individuals
  if (length(selected) < num_individuals) {
    additional <- sample(seq_len(nrow(population)), num_individuals - length(selected), replace = TRUE)
    selected <- c(selected, additional)
  } else if (length(selected) > num_individuals) {
    selected <- sample(selected, num_individuals)
  }
  
  return(population[selected, , drop = FALSE])
}

# Function to generate high-depth sequences by replicating haplotypes
generate_high_depth_sequences <- function(population, depth_per_individual = 20) {
  high_depth_sequences <- list()
  
  for (i in seq_len(nrow(population))) {
    # Replicate each individual’s haplotype pair (A1 + B1, A2 + B2) multiple times to simulate read depth
    for (j in 1:depth_per_individual) {
      high_depth_sequences <- c(high_depth_sequences, paste(population$Hap_A1[i], population$Hap_B1[i], sep = ""))
      high_depth_sequences <- c(high_depth_sequences, paste(population$Hap_A2[i], population$Hap_B2[i], sep = ""))
    }
  }
  
  return(high_depth_sequences)
}

# Generate DNA sequences for output
generate_dna_sequences <- function(population) {
  sequences <- c(
    paste(population$Hap_A1, population$Hap_B1, sep = ""),
    paste(population$Hap_A2, population$Hap_B2, sep = "")
  )
  return(sequences)
}

# Initialize population
population <- initialize_population(num_individuals, gene_length)

# Simulate over generations
for (generation in seq_len(num_generations)) {
  # Reproduction: Create offspring by sampling parents
  offspring <- population[sample(seq_len(nrow(population)), num_individuals, replace = TRUE), ]
  
  # Mutation
  offspring <- mutate_population(offspring, mutation_rate)
  
  # Recombination
  offspring <- recombine(offspring)
  
  # Selection
  offspring <- apply_selection(offspring, "plastic")
  
  # Update population
  population <- offspring
}

# Generate final high-depth DNA sequences by replicating each individual’s haplotypes
final_high_depth_sequences <- generate_high_depth_sequences(population, depth_per_individual = 50)  # Set desired read depth per individual

# Create sequence names
seq_names <- paste0("Individual_", rep(seq_len(num_individuals), each = 2 * 50), "_Hap_", rep(c("1", "2"), each = 50), "_Rep_", rep(1:50, times = num_individuals))

# Convert character vector to DNAStringSet
fasta_sequences <- DNAStringSet(sapply(final_high_depth_sequences, DNAString))

# Assign names to the sequences
names(fasta_sequences) <- seq_names

# Output separate FASTA files for each individual
for (i in 1:num_individuals) {
  # Correctly subset 100 sequences per individual (50 reads for each haplotype)
  individual_sequences <- fasta_sequences[((i - 1) * 100 + 1):(i * 100)]  
  writeXStringSet(individual_sequences, paste0("Individual_", i, "_high_depth.fasta"))
}

# Reference sequence
reference_sequence <- individual_sequences[[1]]

# Convert to a DNAStringSet object
reference_fasta <- DNAStringSet(DNAString(reference_sequence))

# Assign a name to the reference sequence
names(reference_fasta) <- "Individual_1_Hap_1"

# Write the reference genome to a FASTA file
writeXStringSet(reference_fasta, "reference_genome.fasta")

# ----------------------------------------------
# Align each individual’s reads to the reference genome
for (i in 1:num_individuals) {
  # Align the reads of each individual to the reference
  system(paste("bwa index reference_genome.fasta"))  # Index the reference genome
  system(paste("bwa mem reference_genome.fasta Individual_", i, "_high_depth.fasta > Individual_", i, "_aligned_reads.sam", sep = ""))  # Perform the alignment

  # Convert SAM to BAM
  system(paste("samtools view -Sb Individual_", i, "_aligned_reads.sam > Individual_", i, "_aligned_reads.bam", sep = ""))
  
  # Sort and index the BAM file
  system(paste("samtools sort Individual_", i, "_aligned_reads.bam -o Individual_", i, "_aligned_reads_sorted.bam", sep = ""))
  system(paste("samtools index Individual_", i, "_aligned_reads_sorted.bam", sep = ""))
  
  # Call variants from the BAM file
  system(paste("bcftools mpileup -O v -f reference_genome.fasta Individual_", i, "_aligned_reads_sorted.bam | bcftools call -mv -Ov -o Individual_", i, "_raw_variants.vcf", sep = ""))
  
  # **Filter for biallelic SNPs only**
  system(paste("bcftools view -v snps -m2 -M2 Individual_", i, "_raw_variants.vcf > Individual_", i, "_biallelic_variants.vcf", sep = ""))
}

# Print out the first few lines of the VCF files to verify
system("cat Individual_1_variants.vcf")

```

Merge simulated vcf's

```{bash}

# Compress and index each VCF file
for vcf in *.vcf; do
  bgzip -c "$vcf" > "${vcf}.gz"
  tabix -p vcf "${vcf}.gz"
done

# Merge individual VCF files into a single VCF
bcftools merge -Ov -o merged_variants.vcf *biallelic_variants.vcf.gz

```

Second approach

```{r}

# Load necessary library
library(Biostrings)

# Define parameters
total_length <- 100000  # 100 kb region
num_individuals <- 10   # Population size
num_generations <- 50   # Number of generations
alleles <- c("A", "T", "C", "G")  # Possible nucleotides
gene_length <- total_length / 2   # Length for each haplotype
mutation_rate <- 0.001  # Mutation rate per site

# Function to initialize the population with random haplotypes
initialize_population <- function(n, gene_length) {
  population <- data.frame(
    Hap_A1 = replicate(n, paste(sample(alleles, gene_length, replace = TRUE), collapse = "")),
    Hap_A2 = replicate(n, paste(sample(alleles, gene_length, replace = TRUE), collapse = "")),
    Hap_B1 = replicate(n, paste(sample(alleles, gene_length, replace = TRUE), collapse = "")),
    Hap_B2 = replicate(n, paste(sample(alleles, gene_length, replace = TRUE), collapse = ""))
  )
  return(population)
}

# Function to introduce mutations
mutate_haplotype <- function(haplotype, mutation_rate) {
  seq_split <- unlist(strsplit(haplotype, ""))
  mutate_indices <- which(runif(length(seq_split)) < mutation_rate)
  seq_split[mutate_indices] <- sample(alleles, length(mutate_indices), replace = TRUE)
  return(paste(seq_split, collapse = ""))
}

mutate_population <- function(population, mutation_rate) {
  population[] <- lapply(population, function(hap) sapply(hap, mutate_haplotype, mutation_rate))
  return(population)
}

# Function to perform recombination
recombine <- function(population) {
  n <- nrow(population)
  for (i in seq_len(n)) {
    partner <- sample(seq_len(n), 1)
    if (runif(1) < 0.5) {
      temp <- population$Hap_A2[i]
      population$Hap_A2[i] <- population$Hap_A2[partner]
      population$Hap_A2[partner] <- temp
    }
    if (runif(1) < 0.5) {
      temp <- population$Hap_B2[i]
      population$Hap_B2[i] <- population$Hap_B2[partner]
      population$Hap_B2[partner] <- temp
    }
  }
  return(population)
}

# Function to apply selection while maintaining population size
apply_selection <- function(population, selection_criteria) {
  selected <- switch(
    selection_criteria,
    "plastic" = which(substr(population$Hap_A1, 1, 1) == "A" & substr(population$Hap_B1, 1, 1) == "C"),
    "canalized" = which(substr(population$Hap_A1, 1, 1) == "T" & substr(population$Hap_B1, 1, 1) == "G")
  )
  
  # Ensure selected population is exactly num_individuals
  if (length(selected) < num_individuals) {
    additional <- sample(seq_len(nrow(population)), num_individuals - length(selected), replace = TRUE)
    selected <- c(selected, additional)
  } else if (length(selected) > num_individuals) {
    selected <- sample(selected, num_individuals)
  }
  
  return(population[selected, , drop = FALSE])
}

# Generate DNA sequences for output
generate_dna_sequences <- function(population) {
  sequences <- c(
    paste(population$Hap_A1, population$Hap_B1, sep = ""),
    paste(population$Hap_A2, population$Hap_B2, sep = "")
  )
  return(sequences)
}

# Initialize population
population <- initialize_population(num_individuals, gene_length)

# Simulate over generations
for (generation in seq_len(num_generations)) {
  # Reproduction: Create offspring by sampling parents
  offspring <- population[sample(seq_len(nrow(population)), num_individuals, replace = TRUE), ]
  
  # Mutation
  offspring <- mutate_population(offspring, mutation_rate)
  
  # Recombination
  offspring <- recombine(offspring)
  
  # Selection
  offspring <- apply_selection(offspring, "plastic")
  
  # Update population
  population <- offspring
}

# Generate final DNA sequences
final_sequences <- generate_dna_sequences(population)

# Create sequence names
seq_names <- paste0("Individual_", rep(seq_len(num_individuals), each = 2), "_Hap_", c("1", "2"))

# Convert sequences into a Biostrings object and write to FASTA
fasta_sequences <- DNAStringSet(final_sequences)
names(fasta_sequences) <- seq_names
writeXStringSet(fasta_sequences, "simulated_population.fasta")

# Print the first few sequences to verify
head(fasta_sequences)

# Select a haplotype to be used as the reference genome (e.g., Individual_1_Hap_1)
reference_sequence <- final_sequences[1]  # Assuming the first sequence is Individual_1_Hap_1
reference_fasta <- DNAStringSet(reference_sequence)

# Assign a name to the reference sequence
names(reference_fasta) <- "Individual_1_Hap_1"

# Write the reference genome to a FASTA file
writeXStringSet(reference_fasta, "reference_genome.fasta")

# ----------------------------------------------
# Align the simulated sequences to the reference genome

# Assuming you have BWA or another aligner installed on the system
# This step would be done via system calls or external software (BWA, Minimap2, etc.)
# The alignment process would generate a BAM file

system("bwa index reference_genome.fasta")  # Index the reference genome
system("bwa mem reference_genome.fasta simulated_population.fasta > aligned_reads.sam")  # Perform the alignment

# Convert SAM to BAM
system("samtools view -Sb aligned_reads.sam > aligned_reads.bam")

# Sort and index the BAM file
system("samtools sort aligned_reads.bam -o aligned_reads_sorted.bam")
system("samtools index aligned_reads_sorted.bam")

# ----------------------------------------------
# Call variants from the BAM file

# Create a VCF file using the `bcftools` command line tool
system("bcftools mpileup -O v -f reference_genome.fasta aligned_reads_sorted.bam | bcftools call -mv -Ov -o variants.vcf")

# Print out the first few lines of the VCF file to verify
system("head variants.vcf")

```

argweaver code

```{bash}

# Load dependencies
module load miniconda3
conda activate argweaver

# Move directories
cd /hb/home/snbogan/pfe

# Run argweaver on aligned samples
arg-sample -f simulated_population.fasta \
 -o /hb/home/snbogan/pfe/argweaver_sim1 \
 -N 10000 -m 2.5e-8 -r 1.5e-8 \
 --ntimes 20 --maxtime 100e3 -c 1 -n 100 --overwrite

```

Wrangling argweaver samples

```{python}

def extract_newick_trees(file_path, output_path):
    """
    Extracts Newick format trees from a file and saves them to a new file.

    Parameters:
        file_path (str): Path to the input file containing Newick trees.
        output_path (str): Path to save the extracted Newick trees.
    """
    trees = []

    with open(file_path, 'r') as file:
        for line in file:
            if line.startswith('TREE'):
                # Extract the Newick tree part from the line
                parts = line.strip().split('\t')
                if len(parts) > 3:
                    newick_tree = parts[3]
                    trees.append(newick_tree)

    # Save extracted trees to output file
    with open(output_path, 'w') as output_file:
        for tree in trees:
            output_file.write(tree + '\n\n')

# Example usage
input_file = '/hb/home/snbogan/pfe/argweaver_sim1.100.smc'  # Replace with your input file path
output_file = 'newick_trees.txt'  # Replace with your desired output file path
extract_newick_trees(input_file, output_file)

```

Extract specific newick trees from sampled geneologies and create mock allele frequencies for first test

```{bash}

sed -n "5p" newick_trees.txt > gen5.newick

for i in {1..20}; do echo "$RANDOM / 32768" | bc -l; done > freq.txt

```


Revised Fig. 3

Title: Factors shaping selection on phenotypic plasticity with empirical support. Add a little pie chart showing what proportion of studies confirm or reject the model in the top right corner of each panel.

Panel A - y = selection coefficient on plasticity, x = environmental stochasticity -> predictability. Absolute variance goes 3,2,1,0,1,2,3. Logistical (?) curves that increase in slope with greater plasticity. Level of plsticity is based on shade (lighter = more plastic). Cartoon of stochastic variance and predictable variance above.

Panel B - y = selection coefficient, x = organismal condition. Greater plasticity shows stronger slopes, logistically increasing selection with improved condition. Cartoon of organismal condition and energetic availability. Cite appropriate condition-dependence example for plasticity.

Panel C - x axis = phenotypic level of secondary trait, y = selection coefficient on plasticity. Greater plasticity results in more negative selection as secondary trait increases. Above, a cartoon of a famous fitness trade off in phenotypes or in an example of a plasticity tradeoff with the citation.

Panel D - y = selection coefficient, x = genetic correlation, assuming both traits are under positive selection. Secondary trait is held constant, as is its selection coefficient. Should be a negative logistic curve. Greater plasticity increases the slope and intensity of the curve. Cartoon of alleles that have opposing pleiotropic effects on plasticity and a secondary trait.

```{r}

env <- seq(-1, 1, length.out = 300)

# Logistic functions for each plasticity level
logistic <- function(x, lower, upper, slope, midpoint) {
  lower + (upper - lower) / (1 + exp(-slope * (x - midpoint)))
}

curve_p0 <- logistic(env, lower = -0.33, upper = 0.33, slope = 5, midpoint = 0)  # inverted
curve_p1 <- logistic(env, lower = -.67, upper = .67, slope = 5, midpoint = 0.15)
curve_p2 <- logistic(env, lower = -1, upper = 1, slope = 5, midpoint = 0.30)

# Combine into one data frame
panel_a <- bind_rows(
  data.frame(env = (env+1)/2, plasticity = 1, selection = curve_p0),
  data.frame(env = (env+1)/2, plasticity = 2, selection = curve_p1),
  data.frame(env = (env+1)/2, plasticity = 3, selection = curve_p2)
)

panel_a$shade <- factor(panel_a$plasticity)

# Plot
plot_a <- ggplot(panel_a, aes(x = env, y = selection, color = shade, group = shade)) +
  geom_line(size = 2) +
  scale_color_manual(values = c("gray20", "gray50", "gray80")) +
  scale_x_continuous(labels = function(x) abs(x)) +
  labs(
    x = "Environmental predictability",
    y = "Relative fitness",
    subtitle = "Predictable vs. stochastic environments",
    color = "Reaction norm"
  ) +
  theme_classic(base_size = 20) +
  theme(legend.position = c(0.33,0.75))  +
  scale_y_continuous(limits = c(-1,1))

plot_a

cond <- seq(-1, 1, length.out = 300)

curve_p0c <- logistic(cond, lower = -.33, upper = .33, slope = 4, midpoint = 0)
curve_p1c <- logistic(cond, lower = -.67, upper = .67, slope = 4, midpoint = 0)
curve_p2c <- logistic(cond, lower = -1, upper = 1, slope = 4, midpoint = 0)

# Combine into one data frame
panel_b <- bind_rows(
  data.frame(cond = (cond + 1)/2, plasticity = 0, selection = curve_p0c),
  data.frame(cond = (cond + 1)/2, plasticity = 1, selection = curve_p1c),
  data.frame(cond = (cond + 1)/2, plasticity = 2, selection = curve_p2c)
)

panel_b$shade <- factor(panel_b$plasticity)

plot_b <- ggplot(panel_b, aes(x = cond, y = selection, color = shade, group = shade)) +
  geom_line(size = 2) +
  scale_color_manual(values = c("gray20", "gray50", "gray80")) +
  labs(x = "Organismal condition",
       y = "Relative fitness",
       subtitle = "Condition dependence") +
  theme_classic(base_size = 20) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(-1,1))

trait <- - seq(-1, 1, length.out = 300)

curve_p0t <- -logistic(trait, lower = -0.33, upper = 0.33, slope = 5, midpoint = 0)
curve_p1t <- -logistic(trait, lower = -.67, upper = .67, slope = 5, midpoint = 0)
curve_p2t <- -logistic(trait, lower = -1, upper = 1, slope = 5, midpoint = 0)

# Combine into one data frame
panel_c <- bind_rows(
  data.frame(trait = trait + 2, plasticity = 1, selection = curve_p0),
  data.frame(trait = trait + 2, plasticity = 2, selection = curve_p1),
  data.frame(trait = trait + 2, plasticity = 3, selection = curve_p2)
)

panel_c$shade <- factor(panel_c$plasticity)

plot_c <- ggplot(panel_c, aes(x = trait, y = selection, color = shade, group = shade)) +
  geom_line(size = 2) +
  scale_color_manual(values = c("gray20", "gray50", "gray80")) +
  labs(x = "Secondary trait phenotype",
       y = "Relative fitness",
       subtitle = "Fitness tradeoff") +
  theme_classic(base_size = 20) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(-1,1))

cor <- seq(-1, 1, length.out = 300)

curve_p0r <- logistic(cor, lower = 0, upper = 0.33, slope = 5, midpoint = 0)  # inverted
curve_p1r <- logistic(cor, lower = 0, upper = .67, slope = 5, midpoint = 0)
curve_p2r <- logistic(cor, lower = 0, upper = 1, slope = 5, midpoint = 0)

# Combine into one data frame
panel_d <- bind_rows(
  data.frame(cor = cor, plasticity = 1, selection = curve_p0r),
  data.frame(cor = cor, plasticity = 2, selection = curve_p1r),
  data.frame(cor = cor, plasticity = 3, selection = curve_p2r)
)

panel_d$shade <- factor(panel_d$plasticity)

plot_d <- ggplot(panel_d, aes(x = cor, y = selection, color = shade, group = shade)) +
  geom_line(size = 2) +
  scale_color_manual(values = c("gray20", "gray50", "gray80")) +
  labs(x = "Genetic correlation",
       y = "Relative fitness",
       subtitle = "Pleiotropy and linkage") +
  theme_classic(base_size = 20) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(-1,1))

Fig3_rev <- ggarrange(
  plot_a,
  plot_b, 
  plot_c,
  plot_d,
  ncol = 2, nrow = 2, 
  labels = c("A", "B", "C", "D")
)

Fig3_rev

# Export the comparison plot as PNG
png("~/Documents/GitHub/PFE_perspective/Fig3_rev.png", 
    units = "in", width = 12, height = 10, res = 600)
Fig3_rev
dev.off()

```

